### 2023-10-25 abstract.tex
Updates from Overleaf
Fix a bunch of typos

The marked lambda calculus is bidirectionally typed, so localization decisions are systematically predictable based on a local flow of typing information. 
Constraint-based type inference can bring more distant information to bear in discovering inconsistencies but this notoriously complicates error localization. We approach this problem by deploying constraint solving as a type-hole-filling layer atop this gradual bidirectionally typed core. Errors arising from inconsistent unification constraints are localized exclusively to type and expression holes, i.e. the system identifies unfillable holes using a sytem of traced  provenances, rather than localized in an \emph{ad hoc} manner to particular expressions. The user can then interactively shift these errors to particular downstream expressions by selecting from suggested partially consistent type hole fillings, which returns control back to the bidirectional system. We implement this type hole inference system in Hazel. 
\end{abstract}

>>>>

The marked lambda calculus is bidirectionally typed, so localization decisions are systematically predictable based on a local flow of typing information. 
Constraint-based type inference can bring more distant information to bear in discovering inconsistencies but this notoriously complicates error localization. We approach this problem by deploying constraint solving as a type-hole-filling layer atop this gradual bidirectionally typed core. Errors arising from inconsistent unification constraints are localized exclusively to type and expression holes, i.e. the system identifies unfillable holes using a system of traced  provenances, rather than localized in an \emph{ad hoc} manner to particular expressions. The user can then interactively shift these errors to particular downstream expressions by selecting from suggested partially consistent type hole fillings, which returns control back to the bidirectional system. We implement this type hole inference system in Hazel. 
\end{abstract}

### 2023-10-25 calculus.tex
Fix a bunch of typos

The key contribution of this section is the \emph{marked lambda calculus}, a calculus based on the
gradually typed lambda calculus (GTLC) that formalizes the mechanism by which errors like these can
be localized, and how to recover, in all cases, from such errors. 
% In brief, the marked lambda calculus formalizes bidirectional type error localization with gradual
% recovery. 
\Cref{sec:calculus-calculus} introduces the syntax, judgemental structure, guiding metatheory, and
then goes through the rules, organized by syntactic form rather than judgement, to build intuition
about how the various judgements relate to one another \cite{Siek06a}.
All rules and theorems may be found organized by judgement form in the supplementary appendix for
reference, alongside a complete mechanization in the Agda proof assistant \cite{norell:thesis}
(discussed in \cref{sec:calculus-agda}).
We intentionally keep the marked lambda calculus minimal, because it is intended to capture the
essential idea and introduce a general pattern that language designers can employ to created mark
variants of their own type systems.
As an initial example, we consider a surprisingly subtle combination of features as a more
substantial case study: the combination of destructing let expressions with granular type
annotations, in \cref{sec:calculus-let}.
Finally, \cref{sec:calculus-polymorphism} briefly explores how the system might be extended to more
complex judgemental structures, such as parametric polymorphism.

@@

    \judgment{
      \consistent{\TMV_1}{\TMV_1'} \\
      \consistent{\TMV_2}{\TMV_2'} \\
    }{
      \consistent{\TArrow{\TMV_1}{\TMV_2}}{\TArrow{\TMV_1'}{\TMV_2'}}
    }{TCArr}
  \end{mathpar}
  \vspace{-10px}
  \caption{Type consistency.}
  \label{fig:calculus-consistency}
\end{figure}

@@

Unlike numbers and variables, there are explicit synthesis and analysis rules for unmarked lambda
abstractions. This is because expected input and output types are known, and we may verify that the
type annotation and body match them.
%
\begin{mathpar}
  \judgment{
    \ctxSynTypeU{\extendCtx{\ctx}{x}{\TMV}}{\EMV}{\TMV_2}
  }{
    \ctxSynTypeU{\ctx}{\ELam{x}{\TMV_1}{\EMV}}{\TArrow{\TMV_1}{\TMV_2}}
  }{USLam}

@@

    \judgment{ }{
      \matchedArrow{\TArrow{\TMV}{\TMV}}{\TMV}{\TMV}
    }{TMAArr}
  \end{mathpar}
  \vspace{-10px}
  \caption{Matched arrow types.}
  \label{fig:calculus-matched-arrow}
\end{figure}

@@

  \judgment{
    \ctxSynTypeM{\extendCtx{\ctx}{x}{\TMV}}{\ECMV}{\TMV_2}
  }{
    \ctxSynTypeM{\ctx}{\ECLam{x}{\TMV_1}{\ECMV}}{\TArrow{\TMV_1}{\TMV_2}}
  }{MSLam}
\end{mathpar}
%
Similarly, we construct corresponding analytic rules:
%
\begin{mathpar}
  \judgment{
    \matchedArrow{\TMV_3}{\TMV_1}{\TMV_2} \\
    \consistent{\TMV}{\TMV_1} \\\\
    \ctxAnaFixedInto{\extendCtx{\ctx}{x}{\TMV}}{\EMV}{\ECMV}{\TMV_2}
  }{
    \ctxAnaFixedInto{\ctx}{\ELam{x}{\TMV}{\EMV}}{\ECLam{x}{\TMV}{\ECMV}}{\TMV_3}
  }{MKALam1}

@@

Another error arises when $\matchedArrow{\TMV_3}{\TMV_1}{\TMV_2}$, but
$\inconsistent{\TMV}{\TMV_1}$, i.e. the actual type annotation for $x$ is inconsistent with the
expected input type. The \emph{inconsistent ascription mark} $\ECLamInconAsc{x}{\TMV}{\ECMV}$ indicates exactly this
error, and we add a final pair of analytic rules:
%
\begin{mathpar}
  \judgment{
    \matchedArrow{\TMV_3}{\TMV_1}{\TMV_2} \\
    \inconsistent{\TMV}{\TMV_1} \\\\
    \ctxAnaFixedInto{\extendCtx{\ctx}{x}{\TMV_1}}{\EMV}{\ECMV}{\TMV_2}
  }{
    \ctxAnaFixedInto{\ctx}{\ELam{x}{\TMV}{\EMV}}{\ECLamInconAsc{x}{\TMV}{\ECMV}}{\TMV_3}
  }{MKALam3}

@@

\newcommand{\meetsTo}[3]{\ensuremath{\TMeet{#1}{#2} & = & #3}}
\begin{figure}[htbp]
  \raggedright
  \judgbox{\ensuremath{\TMeet{\TMV_1}{\TMV_2}}} is a \emph{partial} metafunction defined as follows:
  %
  \[\begin{array}{rcl}
    \meetsTo{\TUnknown}{\TMV}{\TMV} \\
    \meetsTo{\TMV}{\TUnknown}{\TMV} \\
    \meetsTo{\TNum}{\TNum}{\TNum} \\
    \meetsTo{\TBool}{\TBool}{\TBool} \\
    \meetsTo{(\TArrow{\TMV_1}{\TMV_2})}{(\TArrow{\TMV_1'}{\TMV_2'})}{\TArrow{(\TMeet{\TMV_1}{\TMV_1'})}{(\TMeet{\TMV_2}{\TMV_2'})}}
  \end{array}\]
  \vspace{-10px}
  \caption{Type meet.}
  \label{fig:calculus-type-meet}
\end{figure}

@@

In the case that the subject of the projection does not synthesize a matched product type, we mark with an error, written $\ECProjLSynNonMatchedProd{\ECMV}$:
%
\begin{mathpar}
  \inferrule[MKSProjL2]{
    \ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV} \\
    \notMatchedProd{\TMV}
  }{
    \ctxSynFixedInto{\ctx}{\EProjL{\EMV}}{\ECProjLSynNonMatchedProd{\ECMV}}{\TMV_1}
  }

@@

We have implemented a marking system based on the marked lambda calculus as the foundation for a new version of the Hazel. % TODO citation and explain what in the world Hazel is (reviewer A stated that the lack of this made this section pretty much impossible to understand)
In particular, we have implemented marking for all of the features in the previous section as well as Hazel's $n$-tuples, lists, algebraic datatypes, general pattern matching, strings, and explicit polymorphism (all following the same recipe we developed above).
Hazel is implemented in OCaml and compiles to Javascript via \li{js_of_ocaml} \cite{jsoo} for use in the browser.
The supplemental material contains a pre-built implementation.

>>>>

The key contribution of this section is the \emph{marked lambda calculus}, a calculus based on the
gradually typed lambda calculus (GTLC) that formalizes the mechanism by which errors like these can
be localized, and how to recover, in all cases, from such errors. 
% In brief, the marked lambda calculus formalizes bidirectional type error localization with gradual
% recovery. 
\Cref{sec:calculus-calculus} introduces the syntax, judgemental structure, guiding metatheory, and
then goes through the rules, organized by syntactic form rather than judgement, to build intuition
about how the various judgements relate to one another \cite{Siek06a}.
All rules and theorems may be found organized by judgement form in the supplementary appendix for
reference, alongside a complete mechanization in the Agda proof assistant \cite{norell:thesis}
(discussed in \cref{sec:calculus-agda}).
We intentionally keep the marked lambda calculus minimal, because it is intended to capture the
essential idea and introduce a general pattern that language designers can employ to created marked
variants of their own type systems.
As an initial example, we consider a surprisingly subtle combination of features as a more
substantial case study: the combination of destructing let expressions with granular type
annotations, in \cref{sec:calculus-let}.
Finally, \cref{sec:calculus-polymorphism} briefly explores how the system might be extended to more
complex judgemental structures, such as parametric polymorphism.

@@

    \judgment{
      \consistent{\TMV_1}{\TMV_1'} \\
      \consistent{\TMV_2}{\TMV_2'} \\
    }{
      \consistent{\TArrow{\TMV_1}{\TMV_2}}{\TArrow{\TMV_1'}{\TMV_2'}}
    }{TCArr}

    \inferrule[TCProd]{
      \consistent{\TMV_1}{\TMV_1'} \\
      \consistent{\TMV_2}{\TMV_2'} \\
    }{
      \consistent{\TProd{\TMV_1}{\TMV_2}}{\TProd{\TMV_1'}{\TMV_2'}}
    }
  \end{mathpar}
  \vspace{-10px}
  \caption{Type consistency.}
  \label{fig:calculus-consistency}
\end{figure}

@@

Unlike numbers and variables, there are explicit synthesis and analysis rules for unmarked lambda
abstractions. This is because expected input and output types are known, and we may verify that the
type annotation and body match them.
%
\begin{mathpar}
  \judgment{
    \ctxSynTypeU{\extendCtx{\ctx}{x}{\TMV_1}}{\EMV}{\TMV_2}
  }{
    \ctxSynTypeU{\ctx}{\ELam{x}{\TMV_1}{\EMV}}{\TArrow{\TMV_1}{\TMV_2}}
  }{USLam}

@@

    \judgment{ }{
      \matchedArrow{\TArrow{\TMV_1}{\TMV_2}}{\TMV_1}{\TMV_2}
    }{TMAArr}
  \end{mathpar}
  \vspace{-10px}
  \caption{Matched arrow types.}
  \label{fig:calculus-matched-arrow}
\end{figure}

@@

  \judgment{
    \ctxSynTypeM{\extendCtx{\ctx}{x}{\TMV_1}}{\ECMV}{\TMV_2}
  }{
    \ctxSynTypeM{\ctx}{\ECLam{x}{\TMV_1}{\ECMV}}{\TArrow{\TMV_1}{\TMV_2}}
  }{MSLam}
\end{mathpar}
%
Similarly, we construct corresponding analytic rules:
%
\begin{mathpar}
  \judgment{
    \matchedArrow{\TMV_3}{\TMV_1}{\TMV_2} \\
    \consistent{\TMV}{\TMV_1} \\\\
    \ctxAnaFixedInto{\extendCtx{\ctx}{x}{\TMV}}{\EMV}{\ECMV}{\TMV_2}
  }{
    \ctxAnaFixedInto{\ctx}{\ELam{x}{\TMV}{\EMV}}{\ECLam{x}{\TMV}{\ECMV}}{\TMV_3}
  }{MKALam1}

@@

Another error arises when $\matchedArrow{\TMV_3}{\TMV_1}{\TMV_2}$, but
$\inconsistent{\TMV}{\TMV_1}$, i.e. the actual type annotation for $x$ is inconsistent with the
expected input type. The \emph{inconsistent ascription mark} $\ECLamInconAsc{x}{\TMV}{\ECMV}$ indicates exactly this
error, and we add a final pair of analytic rules:
%
\begin{mathpar}
  \judgment{
    \matchedArrow{\TMV_3}{\TMV_1}{\TMV_2} \\
    \inconsistent{\TMV}{\TMV_1} \\\\
    \ctxAnaFixedInto{\extendCtx{\ctx}{x}{\TMV}}{\EMV}{\ECMV}{\TMV_2}
  }{
    \ctxAnaFixedInto{\ctx}{\ELam{x}{\TMV}{\EMV}}{\ECLamInconAsc{x}{\TMV}{\ECMV}}{\TMV_3}
  }{MKALam3}

@@

\newcommand{\meetsTo}[3]{\ensuremath{\TMeet{#1}{#2} & = & #3}}
\begin{figure}[htbp]
  \raggedright
  \judgbox{\ensuremath{\TMeet{\TMV_1}{\TMV_2}}} is a \emph{partial} metafunction defined as follows:
  %
  \[\begin{array}{rcl}
    \meetsTo{\TUnknown}{\TMV}{\TMV} \\
    \meetsTo{\TMV}{\TUnknown}{\TMV} \\
    \meetsTo{\TNum}{\TNum}{\TNum} \\
    \meetsTo{\TBool}{\TBool}{\TBool} \\
    \meetsTo{(\TArrow{\TMV_1}{\TMV_2})}{(\TArrow{\TMV_1'}{\TMV_2'})}{\TArrow{(\TMeet{\TMV_1}{\TMV_1'})}{(\TMeet{\TMV_2}{\TMV_2'})}} \\
    \meetsTo{(\TProd{\TMV_1}{\TMV_2})}{(\TProd{\TMV_1'}{\TMV_2'})}{\TProd{(\TMeet{\TMV_1}{\TMV_1'})}{(\TMeet{\TMV_2}{\TMV_2'})}}
  \end{array}\]
  \vspace{-10px}
  \caption{Type meet.}
  \label{fig:calculus-type-meet}
\end{figure}

@@

In the case that the subject of the projection does not synthesize a matched product type, we mark with an error, written $\ECProjLSynNonMatchedProd{\ECMV}$:
%
\begin{mathpar}
  \inferrule[MKSProjL2]{
    \ctxSynFixedInto{\ctx}{\EMV}{\ECMV}{\TMV} \\
    \notMatchedProd{\TMV}
  }{
    \ctxSynFixedInto{\ctx}{\EProjL{\EMV}}{\ECProjLSynNonMatchedProd{\ECMV}}{\TUnknown}
  }

@@

We have implemented a marking system based on the marked lambda calculus as the foundation for a new version of Hazel. % TODO citation and explain what in the world Hazel is (reviewer A stated that the lack of this made this section pretty much impossible to understand)
In particular, we have implemented marking for all of the features in the previous section as well as Hazel's $n$-tuples, lists, algebraic datatypes, general pattern matching, strings, and explicit polymorphism (all following the same recipe we developed above).
Hazel is implemented in OCaml and compiles to Javascript via \li{js_of_ocaml} \cite{jsoo} for use in the browser.
The supplemental material contains a pre-built implementation.

### 2023-10-25 intro.tex
Writing, cleanup
Fix a bunch of typos

\Cref{sec:calculus} of this paper develops the first comprehensive {type-theoretic formulation} of type error localization and recovery, called the \emph{marked lambda calculus}.
We bring together three individually well-studied ideas: (1) bidirectional typing, which specifies how type and binding information flows from 
type annotations to expressions \cite{Localinf,BidirTyping}, (2) gradual typing, which offers a principled approach for recovering from missing type information \cite{Siek06a,siek2015refined}, and 
(3) non-empty holes, which function as syntactic membranes marking erroneous terms \cite{HazelnutPOPL}, operationalizing the ``red lines'' that an editor displays.
The marked lambda calculus achieves \emph{total error recovery}, meaning that \emph{every} syntactically well-formed program sketch (a program structure with, optionally, empty holes \cite{solar2013program}) can be marked, i.e. its errors can be simultaneously localized, such that the resulting program has a well-defined type and binding structure.
We establish this and other properties as metatheorems that we mechanize in the Agda proof assistant. 
As we define the calculus, we consider a number of situations where error localization decisions are subtle and conclude by defining some extensions of the intentionally minimal core calculus, e.g. with destructuring patterns and System F-style polymorphism, 
that serve as case studies of the general approach that we hope this paper will be adopted by language designers defining practical type systems.

@@

For example, in the following Hazel expression (ignoring the error on the type hole for the moment), bidirectional error localization 
does not mark any uses of \li{x} as erroneous because a type for \li{x} cannot locally be inferred (the let-bound expression is an empty hole, which has unknown type). However, by gathering constraints from the three uses of \li{x} and attempting to unify, we see that the type hole is unfillable. Rather than arbitrarily privileging one of these uses, as is the case in languages like OCaml (and which has led to the development of complex heuristics, e.g. using machine learning \cite{SeidelBlame}, to avoid misleading programmers), this error is localized, with a \li{!}, to the type hole itself, with the error message providing partially consistent solutions as shown.
\begin{center}
\vspace{-3px}
    % \includegraphics[scale=0.55]{images/intro-thi.png}
    % \includegraphics[scale=0.4]{images/example_CI.png}
    \includegraphics[scale=0.4]{images/figSugg.png}
\end{center}

>>>>

\Cref{sec:calculus} of this paper develops the first comprehensive {type-theoretic formulation} of type error localization and recovery, called the \emph{marked lambda calculus}.
We bring together three individually well-studied ideas: (1) bidirectional typing, which specifies how type and binding information flows from 
type annotations to expressions \cite{Localinf,BidirTyping}, (2) gradual typing, which offers a principled approach for recovering from missing type information \cite{Siek06a,siek2015refined}, and 
(3) non-empty holes, which function as syntactic membranes marking erroneous terms \cite{HazelnutPOPL}, operationalizing the ``red lines'' that an editor displays.
The marked lambda calculus achieves \emph{total error recovery}, meaning that \emph{every} syntactically well-formed program sketch (a program structure with, optionally, empty holes \cite{solar2013program}) can be marked, i.e. its errors can be simultaneously localized, such that the resulting program has a well-defined type and binding structure.
We establish this and other properties as metatheorems that we mechanize in the Agda proof assistant. 
As we define the calculus, we consider a number of situations where error localization decisions are subtle and conclude by defining some extensions of the intentionally minimal core calculus, e.g. with destructuring patterns and System F-style polymorphism, 
that serve as case studies of the general approach that we hope will be adopted by language designers defining practical type systems.

@@

For example, in the following Hazel expression (ignoring the error on the type hole for the moment), bidirectional error localization 
does not mark any uses of \li{x} as erroneous because a type for \li{x} cannot locally be inferred (the let-bound expression is an empty hole, which has unknown type). However, by gathering constraints from the three uses of \li{x} and attempting to unify, we see that the type hole is unfillable. Rather than arbitrarily privileging one of these uses, as is the case in languages like OCaml (and which has led to the development of complex heuristics, e.g. using machine learning \cite{SeidelBlame}, to avoid misleading programmers), this error is localized, with a \li{!}, to the type hole itself, with the error message providing partially consistent solutions as shown.
\\
\begin{center}
    % \includegraphics[scale=0.55]{images/intro-thi.png}
    % \includegraphics[scale=0.4]{images/example_CI.png}
    \includegraphics[scale=0.4]{images/figSugg.png}
\end{center}

### 2023-10-25 let.tex
Fix a bunch of typos

@@

@@

Let synthesis is straightforward: we synthesize the definition's type, then analyze the pattern against it:

@@

The analytic rule is similar, with the final
premise and conclusion changed to analysis.
\\
Attempting an analogous marking rule runs into trouble. We want any type annotations in the pattern to be canonical, which means analyzing the definition to attribute to it any inconsistencies. But we still must analyze the pattern, incorporating the definition's type to produce the context needed by the body. So we do a 'round trip': Analyze the definition against the pattern's type, and then the pattern against the definition's. Since the first analysis establishes consistency, this second analysis is guaranteed to succeed; we only care about the context produced:

>>>>

  \inferrule[USPPair]{
    \ctxSynPatU{\ctx}{\PMV_1}{\TMV_1} \\
    \ctxSynPatU{\ctx}{\PMV_2}{\TMV_2}
  }{
    \ctxSynPatU{\ctx}{\PPair{\PMV_1}{\PMV_2}}{\TProd{\TMV_1}{\TMV_2}}
  }

@@

  \inferrule[UAPPair]{
    \matchedProd{\TMV}{\TMV_1}{\TMV_2} \\
    \ctxAnaPatU{\ctx}{\PMV_1}{\TMV_1}{\ctx_1} \\
    \ctxAnaPatU{\ctx_1}{\PMV_2}{\TMV_2}{\ctx_2}
  }{
    \ctxAnaPatU{\ctx}{\PPair{\PMV_1}{\PMV_2}}{\TMV}{\ctx_2}
  }

@@

Synthesis for let expressions is straightforward: we synthesize the definition's type, then analyze the pattern against it:

@@

The analytic rule is similar, with the final
premise and conclusion changed to analysis.

Attempting an analogous marking rule runs into trouble. We want any type annotations in the pattern
to be canonical, which means analyzing the definition to attribute to it any inconsistencies. But we
still must analyze the pattern, incorporating the definition's type to produce the context needed by
the body. So we do a ``round trip'': analyze the definition against the pattern's type, and then
the pattern against the definition's. Since the first analysis establishes consistency, this second
analysis is guaranteed to succeed; we only care about the context produced:
